---
title: "Final Report "
subtitle: "I-590 Applied Data Science"
author: "Bob Palmer"
date: "March 24, 2019"
output: html_document
---

```{r setup, include=FALSE, warning=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE,warning=FALSE) #turn off echo as a default
library(ggplot2)

library(dplyr)
#library(tidyverse)  # data manipulation
#library(cluster)    # clustering algorithms
#library(factoextra) # clustering algorithms & visualization
library(zoo) # interpolation of missing values
library(imputeTS) # na.mean etc. imputation of missing values in data

options(scipen=999)   #turn off scipen=999 scientific notation or back on scipen=0
```



```{r, echo=FALSE}
# reading in the csv file and data quality remediation aka wrangling
rawdata <- read.csv("gapminder.csv", stringsAsFactors = FALSE) # keeps strings, not factors
data <- rawdata
data$population <- as.numeric(gsub(",", "", data$population)) #remove commas from the thousands indicator in numbers
data$population <- na.locf(data$population) #last observation before NA or a string of NA is used to replace na
data$income <- na.mean(data$income) # replace missing incomes with means

uniqueCountry <- length(unique(data$Country))
missing <- rawdata[which(is.na(rawdata[,5])),]
uniqueMissing <- length(unique(missing$Country))
```

```{r}
# Overall narrative summary of data:
# how many observations 2pts
# how many variables 2pts
# type of variables 2pts
# how disperse is your data (e.g price), range of variables (e.g. years) 2pts
# data wrangling 2pts
# preprocessing steps (e.g. - normalization, log scale). If no preprocessing - explain why
# Note: in gapminder population is calculated every ten years. Thus the same # population is applied to every year during the 10 year period. Consider wrangling your data before the analysis, do not treat these values as zero.
# Explore data 60 pts - provide insights from data
```

## I. Description of the Data. 


This report is based on the gapminder data file available from Gapminder.org. There are `r nrow(data)` observations in the file, covering `r uniqueCountry` unique countries. The data provides `r length(data) ` variables: Country, Year, life (expectancy) , population, GDP income per capita, and region. The dispersion of the data varies;  for most countries the years are given from 1800 to 2015, but for some for some countries the years are from 1970 to 2015. 

Because for many countries population is given only once every ten years (presumably a decennial cencus) we have interpolated the missing values from the previous cencus until the next cencus occurs. 
We can also see that there were `r sum(is.na(rawdata$income)) ` missing values in the income column; those missing income values area from `r uniqueMissing ` countries. We have interpolated the missing values with the mean of the income column.  

Countries are identified by five regions: South Asia, Middle East & North Africa, Sub-Saharan Africa, Europe & Central Asia, and America. We can see that the country and population variables are character strings, year and income are integers, and The life variable is a numeric. Commas separating thousands digits have been removed from some observations of the population data. 


##II. Questions to Ask 

Interesting questions that we can ask about this data set include:

a. Which region has the highest median life expectancy and which has the lowest?

b. Which region has the highest median per capita GDP income and which has the lowest?

c. How does life expectancy change over time for the regions?

d. How does income change over time for the regions?

e. Is income a predictor of life expectancy?

f. Is  region a predictor of life expectancy?

##III. Exploration of the Data 

```{r}
# Explore data 60 pts - provide insights from data
# 
# Exploratory plots 20 pts
# histogram, correlation plot, box plot (your choice of meaningful plots)

# Cluster analysis 20 pts
# Specify explicitly the choice of # clusters using one of the learned methods (elbow, silhouette etc) 5 pts
# Visualize cluster 10 pts
# Summarize the observed pattern (if any) 5pts

# Regression analysis 20 pts
# Specify the formula - what is your dependent variable=outcome (DV) and what are your independent variables/features (IV) 2 pts
# Provide the regression summary table 3 pts
# Explain the summary table 5 pts
# Visualize 4 diagnostic plots 5 pts
# Explain how well your model fit the assumptions based on those diagnostic plots 5 pts
# Note: Provide a full explanation/description for each figure. 
# 
# Subtracting 1pt for not providing explanation for a figure
# 
# Subtracting 1pt for printing data tables. Do not include head functions, do not print a portion of dataframe - it is only needed for your exploration but not for report. Consider simply providing a descriptive information
# 
# Visualization 5 pts (overall)
# 
# Legends 1pt
# Title 1pt
# Axis titles 1pt
# Axis values 1pt
# Plots are readable (for example, avoid colors that are hard to distinguished, it is a good practice to know if your colors are appropriate for color blindness) 1pt
```


An interesting statistic in the summary of the data is that the mean life expectancy for all humans in all countries in the world for all of the years sampled is 42.88 years. 



```{r}
#- provide insights from data.-60 pts
#- exploratory plots (histogram, correlation plot, box plot)  - 20
```

Here is a histogram of the incomes for all countries for all years in the data set: 

```{r}
df <- select(filter(data, income <= 100000),c(income,population))
Income <- df$income
hist.data = hist(Income, plot=F)

plot(hist.data, col = 'blue' )

```

It seems clear from this that income per capita for all countires for all years in the data is definitely not in a gaussian (normal) distribution, but somewhat predictably it looks to be an inverse Pareto distribution, showing the predominance of low incomes in a very large part of the population, and a long thin tail of high incomes. The largest bin of income is at or below 5,000 international dollars. This sharp fall-off of income still makes this hard to visualize, so let's look at the frequency counts in a logarithmic transformation (in this case, log 2).

```{r}
df <- select(filter(data, income <= 100000),c(income,population))
Income <- df$income
hist.data = hist(Income, plot=F)
hist.data$counts = log(hist.data$counts, 2)
plot(hist.data, col = 'blue' )

```

The fact that there are many more poor countries seems intuitive, but later we will see if that changes over the years of the study.   

####Question a) Which region has the highest median life expectancy and which has the lowest?

To answer this question, We explore the data graphically, by aggregating the median life expectancy for each region, and then showing the regions on a box chart. From this chart we can see that the Middle East & North Africa, South Asia and Sub-Saharan Africa regions all have a median life expectancy in the mid 32 years. Europe & Central Asia region has the highest median life expectancy of roughly 41 and three quarter years. (The median values are labeled on the chart in red.) We see that the first quartile of life expectancy for Europe is actually higher than the median (second quartile) of all of the other regions -- in other words the whole interquartile range of Europe is above the median of all the other regions.

```{r, echo=FALSE}
#First, we calculate the region mean life expectancy with aggregate, so we can print it on the boxplot:

medianLife <- aggregate(life ~  region, data, median)

```

```{r, echo=FALSE, fig.height=6, fig.width=8}



ggplot(data, aes(x = region, y = life)) + 
  xlab('') +
  ylab('Life Expectancy') +
  theme(axis.title.x=element_text(angle=0, color='black', size = 16), axis.title.y=element_text(angle=90, color='black', face='bold', size=14)) +
  geom_boxplot(outlier.colour = "grey") + 
  theme(axis.text.x = element_text(angle = 60, size = 12, color = "blue", hjust = 1)) +    theme(axis.text.y = element_text(angle = 90, size = 16, color = "black", hjust = 1)) +   geom_text(data = medianLife, aes(label = round(life, digits = 2), y = life + 3), size = 4, colour = "red")
  
```

####Question b) Which region has the highest median per capita GDP income and which has the lowest?

To answer this, We again build a boxplot by region, by aggregating the median income for each region, and again, the medians are labeled in red. (The extreme outliers in income have been removed for clarity, but the interquartile ranges were not affected by the method.) 

We can see that Europe & Central Asia again is the leader with the highest median income. Sub-Saharan Africa is the lowest median income. 



```{r, echo=FALSE, fig.height=6, fig.width=8}
#First, we calculate the region mean income with aggregate, so we can print it on the boxplot:
medianIncome <- aggregate(income ~ region, data, median)

ggplot(data, aes(x = region, y = income)) + 
  xlab('') +
  ylab('Income Per Capita') +
  theme(axis.title.x=element_text(angle=0, color='black', size = 16), axis.title.y=element_text(angle=90, color='black', face='bold', size=14)) +
  geom_boxplot(outlier.shape = NA) +
  coord_cartesian(ylim = c(0, 15000)) +
    theme(axis.text.x = element_text(angle = 60, size = 12, color = "blue", hjust = 1)) +      theme(axis.text.y = element_text(angle = 90, size = 16, color = "black", hjust = 1)) +   geom_text(data = medianIncome, aes(label = income, y = income + 600), size = 4, colour = "red") 
  
  
```

From the last two figures, we may hypothesize that there is a correlation between a region's per capita income and life expectancy. 

####Question c) How does life expectancy change over time for the regions?

From the scatter plots below, we clearly see that beginning around the 1925-1950, all of the regions show a general trend of increasing life expectancy. We could hypothesize that this is due to increasing nutrition (perhaps driven by artificial nitrogen fertilizers introduced during that period) and increases in the efficacy of medical treatments.


```{r, echo=FALSE, fig.height=6, fig.width=10}

#to see the life expectancy over time across different regions
ggplot(data) + 
  geom_point(aes(x=Year, y=life), alpha = 1/10) +
  xlab('Year') +
  ylab('Life Expectancy') +
  theme(axis.title.x=element_text(angle=0, color='black', size = 14),   axis.title.y=element_text(angle=90, color='black', face='bold', size=14)) +
  facet_wrap(~region) + 
  theme(strip.text.x = element_text(size = 12, colour = "blue"), axis.text=element_text(size=12), axis.title=element_text(size=12,face="bold")) 
```

#### Question d) How does income change over time for the regions?

We can see some interesting changes; East Asia and the Pacific had some marked increases in income in the post WWII era (perhaps driven by the resergence in manufacturing in Asian countries). We can also see an explosion in income in the Middle East after the discovery and exploitation of petroleum that occurred in the post-war time period. 

```{r, echo=FALSE, fig.height=6, fig.width=10}
#to see the income over time across different regions
ggplot(data) + 
  geom_point(aes(x=Year, y=income), alpha = 1/10) +
  xlab('Year') +
  ylab('Income per Capita') +
  theme(axis.title.x=element_text(angle=0, color='black', size = 14),   axis.title.y=element_text(angle=90, color='black', face='bold', size=14)) +
  facet_wrap(~region) + 
  theme(strip.text.x = element_text(size = 12, colour = "blue"), axis.text=element_text(size=12), axis.title=element_text(size=12,face="bold")) 
```

####Question e) Is income a predictor of life expectancy?

#### Linear Regression Analysis

```{r}
# Regression analysis 20 pts
# Specify the formula - what is your dependent variable=outcome (DV) and what are your independent variables/features (IV) 2 pts
# Provide the regression summary table 3 pts
# Explain the summary table 5 pts
# Visualize 4 diagnostic plots 5 pts
# Explain how well your model fit the assumptions based on those diagnostic plots 5 pts
# Note: Provide a full explanation/description for each figure. 
# 
# Subtracting 1pt for not providing explanation for a figure
```


To examine question e) Is income a predictor of life expectancy?, let's do a linear regression analysis on the data. Simple linear regression is a statistical method that allows us to summarize and study relationships between two quantitative variables; in this case the predictor variable (also known as independent variable) is income (x axis), and the response variable is life expectancy (y axis).

In figure below, we plot the raw data of income and life expectancy for the year 2000, and we add a linear model (regression) line. 

```{r echo=FALSE, fig.height=6, fig.width=10}

#select only a specific year of data
income <- subset(data, Year == 2000, select = c("income","life"))

# Simple Scatterplot

plot(life ~ income, data= income, col = 'black', pch = 20, xlab = 'Income', ylab = 'Life Expectancy')

fit = lm(life ~ income, data = income)
abline(fit, col = 'red', lwd = 3)

```

Here is the regression summary table:
```{r}
summary(fit)
```

Multiple R-squared tells us how much of the variablity of the data is expalined by our model. Theoretically, it is the ratio between regression sum of square and total sum of square. It tells us the percentage of the life expectancy variablity which is expalined by the model.

The F-statistic is for testing the model fitness. The F test tells us whether our model assumptions are correct or not. Higher p-value indicates lack of fit. If higher than threshold (0.05 or 0.1) then model should not be accepted. Our F-stastic is very high in this case indicating a poor fit for our linear model. 

From just looking at the the slope of the regression line above we might guess that there is a linear relationship between income and life expectancy, but lets further check if our linear model is valid. We can make a plot of the  residuals vs. predicted values.

Let's look at the distrbution of our residual errors to see if they appear to be in a normal distribution, which would indicate that our model is a good predictor of values for our data. 

```{r}
hist(resid(fit), main = 'Histogram of Residuals of the Linear Model', col = 'blue')
```


Well, the histogram above certainly does not appear to show us a normal distribution. 

A normal Quantile-Quantile plot is another way to tell us whether our errors are in a normal distribution. If they are the plot would follow the line closely. (see below, upper right chart)  Since it does not, we can infer that the residuals are not in a normal distribution, indicating that our model is not very appropriate. 

A Scale-Location Chart (see below, lower left chart) also known as a Spread-Location plot shows us whether residuals are spread equally along the ranges of predictors, in order to check the assumption of equal variance (homoscedasticity). It's good if you see a horizontal line with equally (randomly) spread points. That is clearly not the case here. 

The Residuals vs. Leverage plot (bottom right, below) helps us identify influential data points (like outliers) in our model. The points we're looking for(or not looking for) are values in the upper right or lower right corners, which are outside the red dashed Cook's distance line. These are points that would be influential in the model and removing them would likely alter the regression results.

```{r echo=FALSE, fig.height=8, fig.width=10}
#four diagnostic charts for the linear model
par(mfrow=c(2,2))
plot(fit)
```


####Question f) Is region a predictor of life expectancy?

If region is a predictor of life expectancy, then we could assume that a cluster analysis of the metrics (population, income, life expectancy) of each country would group them in clusters that are similar to their geograpic regions. 

###Cluster Analysis

```{r}
#- cluster analysis - 20

#Note: If you do not describe your analysis, if it will not count toward points. For example, if you #create a histogram but do not provide any explanation of the figure -  it will not be included
```


```{r message=FALSE, warning=FALSE}
library(tidyverse)  # data manipulation
library(cluster)    # clustering algorithms
library(factoextra) # clustering algorithms & visualization
```



Cluster Analysis is a form of exploratory data analysis (EDA) where observations are divided into meaningful groups that share common characteristics (features).

```{r}
###Read in dataset for clustering
df <- read.csv('gapminder.csv')

df <- data[df$Year == '2015',] # filter the data df for only one year data, 

df$population <- as.numeric(gsub(",", "", df$population)) #remove commas from the thousands indicator in numbers

df <- na.omit(df)#Remove missing values

rownames(df) <- df$Country #The industry name will be on the cluster plot instead of row numbers.

dfXindia <- subset(df, Country!="India") # remove india from dataset

dfXindia <- subset(dfXindia, Country!="China") # remove china from dataset

df <- df[, c(3,4,5)] #select only life, population, and income
dfXindia <- dfXindia[, c(3,4,5)] #select only life, population, and income

dat <- scale(df)
datXindia <- scale(dfXindia)

```



```{r}
#function for plotting elbow
wss <- function(k) {
kmeans(df, k, nstart = 10)$tot.withinss
}

k.values <- 1:15

wss_values <- map_dbl(k.values, wss)
```

In order to decide how many clusters to use in our analysis, we can use a so-called elbow chart, which measures the distances differentiating the members of the clusters for different numbers of clusters. 

```{r}
#visualizing the elbow chart
plot(k.values, wss_values, type = "b", pch = 19, frame = FALSE, xlab = "Num of Clusters", ylab="Tot within-clust sum-of-squares")
```

We see that three is a good "elbow point" for selecting the number of clusters to use in our analysis.

However, in the figure below, we see that two outlier countries, India and China are in their own cluster alone. This could be because India and china have an order of magnitude higher population than all other countries in the data set. 


```{r, echo=FALSE, fig.height=8, fig.width=10, warning=FALSE}
#visualiziation
k2 <- kmeans(dat, centers = 3, nstart = 25)
visual<- fviz_cluster(k2, data = dat)
visual
```
So let's see the clusters without the two outlier countries: 



```{r, echo=FALSE, fig.height=8, fig.width=10, warning=FALSE}
# visualization ex-India
k2Xindia <- kmeans(datXindia, centers = 3, nstart = 25)
visualXindia<- fviz_cluster(k2Xindia, data = datXindia)
visualXindia
```

We see that the countries are clustered in such a way that they do not reflect their geographic regions in the clusters. This would not give us enough evidence to support the hypothesis that region is determinitive of life expectancy. 

##IV. Findings 
```{r}
# make sure to provide answers to questions you have generated at the beginning. Write a brief #summary of your answers in your report. You are not expected to have a comprehensive research in #this assignment. 10pts

#6. Use of RMarkdown or Python notebook - 5 pts [proper syntax, documentation, output] Use code #chunks and also in-line #code, header, fonts, title etc
```


To review, here are the six questions we sought to answer with this report. 

 <span style="color:blue">a. Which region has the highest median life expectancy and which has the lowest?</span>

Using the box plots, we discovered that Europe & Central Asia has the hightest median life expectancy, and Middle East & North Africa has the lowest. 

<span style="color:blue">b. Which region has the highest median per capita GDP income and which has the lowest?</span>

We found that Europe & Central Asia has the hightest median income, and Sub-Saharan Africa has the lowest.

<span style="color:blue">c. How does life expectancy change over time for the regions?</span>

The faceted scatter plots for the regions indicated that ll of the regions show a general trend of increasing life expectancy. We could hypothesize that this is due to increasing nutrition (perhaps driven by artificial nitrogen fertilizers introduced during that period) and increases in the efficacy of medical treatments. Data could be found for those observations and that analysis could be the topic of future analysis. 

<span style="color:blue">d. How does income change over time for the regions?</span>

All of the regions had increases in income over the years recorded. East Asia & Pacific had some marked increases in income in the post WWII era (perhaps driven by the resergence in manufacturing in Asian countries). We can also see an explosion in income in the Middle East after the discovery and exploitation of petroleum that occurred in the post-war time period. South Asia and Sub-Saharan Africa regionis did not indicate as large of an increase.

<span style="color:blue">e. Is income a predictor of life expectancy?</span>

Our linear regression analysis was not conclusive in showing that as income increased so did life expectancy. 

<span style="color:blue">f. Is  region a predictor of life expectancy?</span>

Our hypothesis was that if region where a predictor of life expectancy, then we could assume that a cluster analysis of the metrics (population, income, life expectancy) of each country would group them in clusters that are similar to their geograpic regions.  But, this clustering was not upheld by the results. Therefore we do not have sufficient evidence from the data to say that region is determinative of life expectancy. 








